Exploitation Techniques and Mitigations

### 2016-01-16 docs/paper.tex
Paper: Finalize draft of background knowledge

@@

\maketitle

% \section*{Abstract}
% \label{sec:abstract}

% \tableofcontents

% \newpage

@@

@@

@@

>>>>

\newminted{nasm}{fontsize=\scriptsize,frame=leftline,framesep=2mm,linenos}

@@

% \maketitle
%
% \section*{Abstract}
% \label{sec:abstract}
%
% \setcounter{tocdepth}{1}
% \tableofcontents
%
% \newpage

@@

\begin{figure}[htpb]
    \centering
    \includegraphics[width=\textwidth]{gfx/x86_registers.pdf}
    \caption{Register overview including \SI{64}{\bit} extension}
    \label{fig:registers}
\end{figure}

\Cref{fig:registers} (from
Wikipedia\footnote{\url{https://en.wikipedia.org/w/index.php?title=X86&oldid=696308590\#/media/File:Table_of_x86_Registers_svg.svg}})
shows an overview of registers available on the x86 platform. While there are
dedicated registers for floating pointer operations and also special registers
which hardware protection (segment registers) we will only focus on nine most
commonly used registers.

\begin{minipage}[t]{0.5\textwidth}
    \begin{description}
        \item[\texttt{EAX}] Accumulator Register
        \item[\texttt{EBX}] Base Register
        \item[\texttt{ECX}] Counter Register
        \item[\texttt{EDX}] Data Register
        \item[\texttt{ESI}] Source Index
        \item[\texttt{EDI}] Destination Index
        \item[\texttt{EBP}] Base Pointer
        \item[\texttt{ESP}] Stack Pointer
        \item[\texttt{EIP}] Instruction Pointer
    \end{description}
\end{minipage}
\begin{minipage}[t]{0.45\textwidth}
    \begin{figure}[H]
        \includegraphics[width=\textwidth]{gfx/single_register.png}
        \caption{Addressing specific parts of a register including \SI{64}{\bit} extension}
        \label{fig:single_register}
    \end{figure}
\end{minipage}
\bigskip

The instruction pointer \texttt{EIP} points to the next instruction located in
memory which is going to be executed on the cycle. Stack pointer \texttt{ESP} and base
pointer \texttt{EBP} are used for stack management which is vital to call and
return from multiple functions properly. The remaining six registers are used
for computation and passing arguments for system calls. Their values can either
be interpreted as integers or pointers.

Note that these registers can be addressed partially allowing one to write only
to the lower \SI{16}{bit} for example as displayed in
\cref{fig:single_register} taken from \textit{null
programm}\footnote{\url{http://nullprogram.com/img/x86/register.png} on
December of 2015}.

The CPU comes with protection mechanisms which allows the operating system
kernel to limit the privileges of other processes. This mechanism is known as
\textit{protection rings} (Ring 0 -- Ring 3). The kernel runs \emph{in} Ring 0
(most privileged) and switches to Ring 3 (least privileged) when a normal
process is scheduled. A system call has to be made by the process if it needs
something which goes beyond its scope. The kernel takes over, deals with the
request and returns execution back to the process. This is known as
\textit{context switch} and switching between Rings happens along the way.

\subsection{System Calls}

As already mentioned in the previous paragraph, a process only has limited
capabilities and the kernel has to take over to fulfill certain (more
privileged) operations. The operating system's documentation tells you which
system calls are available (on which platform) and what additional parameters
they require. Let us illustrate this with an example: On x86 the Linux system
number 4 (starting from 0) is the \texttt{sys\_write} system call which writes
data to a file descriptor. It takes three arguments, the file descriptor to
write to, a pointer to the start of the data which should be written and the
length of the data. The number of the system call together with these three
parameters are placed in the \texttt{EAX}, \texttt{EBX}, \texttt{ECX},
\texttt{EDX} respectively. To invoke the system call issue following
instruction:

\begin{nasmcode*}{autogobble,xleftmargin=1cm,linenos=false}
    int     0x80
\end{nasmcode*}

Nowadays you may encounter a different mechanism for system calls, the Virtual
Dynamic Shared Objects (vDSO) mechanism. This goes beyond our scope here, we
will use the previously mentioned mechanism in our exploits. If interested, you
may want to look at the related man
page\footnote{\url{http://man7.org/linux/man-pages/man7/vdso.7.html}}.

@@

Physical memory is managed by the operation system kernel by utilising the
Memory Management Unit (MMU). Each process' address space is virtualized and
memory operations are translated on-the-fly by the MMU. The physical memory is
segmented into \textit{pages} (typically \SI{4}{\kibi\byte} in size) and each
page can be mapped into the virtual address space of one or more (shared)
processes.~\cite[pp.~400]{unix_interals}

The main parts located inside the (virtual) address space of a process are the
executable itself with its \texttt{.text} and \texttt{.data} section, the heap
used for dynamic data, the stack used for local variables and function calling
and used libraries.

\subsection{Endianness}

\begin{figure}[t]
    \centering
    \includegraphics[width=0.35\textwidth]{gfx/little_endian.pdf}
    \caption{Placement of bytes in memory in little-endian}
    \label{fig:little_endian}
\end{figure}

Endianness refers to the byte order used when storing data in memory (or
transmitting it over the network). \Cref{fig:little_endian} (from
Wikipedia\footnote{\url{https://en.wikipedia.org/w/index.php?title=Endianness&oldid=696417697\#/media/File:Little-Endian.svg}})
illustrates that the least significant byte of a word is placed at the lower
memory address and successive bytes are placed as the memory address increases.

\subsection{Calling Convention}

A calling convention defines how function calls should be implemented. What
calling convention is used depends on the platform, toolchain and (compiler)
settings. Let us exhibit what the convention defines and what convention we are
using.

\begin{minipage}[t]{0.47\textwidth}
    Defines:
    \begin{itemize}
        \item Where to place arguments
        \item Where to place return value
        \item Where to place return address
        \item Who prepares the stack
        \item Who saves which register
        \item Who cleans up\\
            (caller or callee)
    \end{itemize}
\end{minipage}
\begin{minipage}[t]{0.47\textwidth}
    C Declaration (cdecl):
    \begin{itemize}
        \item Arguments on stack (reverse order)\\
            stack aligned to \SI{16}{\byte} boundary
        \item Return via register (\texttt{EAX} / \texttt{ST0})
        \item \texttt{EAX}, \texttt{ECX}, \texttt{EDX} saved by the caller\\
            rest saved by the callee
        \item On stack:\\
            old instruction pointer (\texttt{IP})\\
            old base pointer (\texttt{BP})
        \item Caller does the cleanup
    \end{itemize}
\end{minipage}

@@

\section{Format String Exploits}

\section{Buffer Overflow}

\section{Shell Code}

\section{Data Execution Prevention (DEP)}

\section{Return Oriented Programming (ROP)}

\section{Address Space Layout Randomization (ASLR)}

\section{Stack Cookies}

\section{Control Flow Integrity (CFI)}

\section{Other Architectures}

\section{Conclusion}
